---
title: "session 1 exercise solutions"
author: "Tim Riffe"
date: "2024-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the data

I got these urls from the _raw_ button of the respective files on github:

```{r, nessage = FALSE}
library(tidyverse)
library(janitor)
B <- read_csv("https://github.com/timriffe/BSSD2024Module2/raw/master/data/ES_B.csv.gz",
              show_col_types = FALSE)
P <- read_csv("https://github.com/timriffe/BSSD2024Module2/raw/master/data/ES_P.csv.gz",
              show_col_types = FALSE)
```

Calculate exposure; note we have `Female1` and `Female2`, these are already year beginning and year end estimates, already side-by side, so this is good. 
```{r}
E <-
  P |> 
  clean_names() |> 
  select(year, age, female1, female2) |> 
  mutate(exposure = (female1 + female2) / 2) |> 
  select(!starts_with("fem"))
```

1. calculate asfr
```{r}
Fx <- 
  B |> 
  clean_names() |> 
  select(year, age, births = total) |> 
  left_join(E, by = join_by(year, age)) |> 
  mutate(Fx = births / exposure)
```

Why not 1975? The deviation in TFR is indeed small; this is coming from the HMD documentation, where we see that there was a universe definition change in that year, which stifles exposure calculations. The HMD has a slick way to account for this, but we don't have the info handy that we'd need.

```{r}
TFR <- 
  Fx |> 
  group_by(year) |> 
  summarize(TFR = sum(Fx)) 
TFR |> 
  ggplot(aes(x = year, y = TFR)) +
  geom_line() +
  geom_point(data = TFR |> filter(year == 1975), color = "red", size = 2)
```

2. plot asfr for 1950 and 2014

```{r}
Fx |> 
  filter(year %in% c(1950,2014)) |> 
  ggplot(aes(x= age, y = Fx, color = as.factor(year))) +
  geom_line()
```

3 and 4: Done above to investigate why 1975 is supposedly problematic

5. and 6. calculate and plot MAB

```{r}
MAB <-
  Fx |> 
  group_by(year) |> 
  summarize(MAB = sum(Fx * age) / sum(Fx) + .5)

MAB |> 
  ggplot(aes(x = year, y = MAB)) +
  geom_line() +
  theme_minimal()
```

Curious how MAB and TFR move together?

```{r}
MAB |> 
  left_join(TFR,by=join_by(year)) |> 
  ggplot(aes(x = TFR, y = MAB, color = year)) +
  geom_point() +
  theme_minimal()
```

