---
title: "session 2 notes"
author: "Tim Riffe"
date: "2024-07-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# grab the data, load tidyverse

```{r}
library(tidyverse)
ES <- read_csv("https://github.com/timriffe/BSSD2024Module2/raw/master/data/ES_mort.csv.gz",
               show_col_types = FALSE)
# ES <- read_csv("data/ES_mort.csv.gz")
```
The second line is for reading in the file locally, where I emphasize that we should always remember to keep the path short, and relative to our current project location.

If your internet is super slow, you can increase the timeout for downloading like so:
```{r}
options(timeout = 1000)
```
This will let your machine keep trying for longer to download a file.

## Compare mortality rates

```{r}
ES |> 
  filter(year %in% c(1930,2019),
         sex == "female") |> 
  ggplot(aes(x = age, y = mx, color = as.factor(year))) +
  geom_line() +
  scale_y_log10()
```

```{r}
ES |> 
  filter(year %in% c(1930,2019),
         sex == "female",
         age < 90) |> 
  select(year, age, mx) |> 
  # year moves to columns here:
  pivot_wider(names_from = year, values_from = mx) |> 
  mutate(mx_diff = `1930` - `2019`) |> 
  ggplot(aes(x = age, y = mx_diff)) +
  geom_line()
```

# $q_x$ conditional death probabilities

$$ q_x = \frac{m_x}{1 + (1 - a_x)m_x} $$

So, the first thing we want is to transform rates to probabilities, behold the above formula. Where on earth is it from? This I encourage you to simply look up whenever you need it rather than memorizing it. I would look it up from here <https://www.mortality.org/File/GetDocument/Public/Docs/MethodsProtocolV6.pdf> or the Preston book whenever I need it. But, look at that tiny variable in the denominator $a_x$: that has to come from somewhere... 

NOTE: annotate this
```{r}
mx_to_ax_simple <- function(mx, age, n){
  tibble(mx,age,n) |> 
    mutate(ax = case_when(
      age == 0 & mx < .02012 ~ .14916 - 2.02536 * mx,
      age == 0 & mx < .07599 ~ 0.037495 + 3.57055 * mx,
      age == 0 & mx >= .07599 ~ 0.30663,
      age == max(age) & mx == 0 ~ .5,
      age == max(age) ~ 1 / mx,
      TRUE ~ n / 2)) |> 
    pull(ax)
}
```

Now we can implement the transformation to $q_x$

We have more details to attend to in this function!
```{r}
mx_to_qx <- function(mx, ax){
  qx <- mx / (1 + (1 - ax) * mx)
  # NEW I
  N <- length(qx)
  qx[N] <- 1
  # NEW II
  qx[qx > 1] <- 1
  return(qx)
}
```

# survivorship: $l_x$

$$ \ell_x = \prod _{t=0}^{x-1} (1 - q_t) \quad \forall x>0$$
$$ \ell_0 = 1 $$

Notes: the `cumprod()` part of this transformation is the most important part.
```{r}
qx_to_lx <- function(qx, radix = 1){
  N  <- length(qx)
  lx <- c(1, cumprod(1 - qx[-N]))
  lx <- lx * radix
  return(lx)
}

# usage:

```
This is the first glance at LE that was have: it's the area under this curve, but we'd want to do a better job integrating that area.


# deaths distribution $d_x$

$$ d_x = \ell_x \cdot q_x $$

Here, I went on a long tangent about the modal age at death, which is great because 
1. it's more stable than $e_0$ (life expectancy at birth)
2. the data requirements for seeings it are far less strict, which means
3. you can get a time series of COHORT modal ages at death even

## $L_x$ is the person-years lived in each age group

$$ L_x = \ell_x - (1 - a_x)d_x \quad \forall x <\omega$$
$$ L_\omega = d_\omega \cdot a_\omega $$

```{r}
lxdx_to_Lx <- function(lx,dx,ax){
  Lx    <- lx - (1 - ax) * dx
  N     <- length(lx)
  Lx[N] <- ax[N] * lx[N]
  return(Lx)
}
```

# Tx Total years lived above x

$T_x$ is an instrumental lifetable column, needed here and there, but not something you look at for demographic insight.

$$ T_x = \sum_x^\omega L_x$$
```{r}
Lx_to_Tx <- function(Lx){
  Tx <- 
    Lx |> 
    rev() |> 
    cumsum() |> 
    rev() 
  return(Tx)
}
```

Now we're ready to calculate full lifetables
$$ e_x = \frac{T_x}{\ell_x}$$

```{r}
LT <- 
 ES |> 
   group_by(sex, year) |> 
   mutate(n = rep(1, n()),
          # add note here
          mx = if_else(is.na(mx), .5, mx),
          ax = mx_to_ax_simple(mx, age, n),
          qx = mx_to_qx(mx, ax),
          lx = qx_to_lx(qx),
          dx = lx * qx,
          Lx = lxdx_to_Lx(lx, dx, ax),
          Tx = Lx_to_Tx(Lx),
          ex = Tx / lx) 
```


# Exercises

1. make plot of $e_0$ over time

```{r}
LT |> 
  filter(age == 0) |> 
  ggplot(aes(x = year, y = ex, color = sex)) +
  geom_line() +
  theme_minimal()
```


2. make plot of $M$ (modal age at death), be sure to filter out ages < 10 for this.

```{r}
LT |> 
  filter(age > 10) |> 
  group_by(sex, year) |> 
  summarize(M = age[dx == max(dx)] + ax[dx == max(dx)]) |> 
  ggplot(aes(x = year, y = M, color = sex)) +
  geom_step() +
  theme_minimal()
```

3. find a mortality shock year from (1), and compare it's $\ell_x$ and $d_x$ with those of a typical year.

```{r, warning = FALSE}
LT |> 
  filter(year %in% c(1917,1918),
         sex == "female",
         age > 10) |> 
  ggplot(aes(x = age, y = dx, color = as.factor(year))) +
  geom_point() +
  geom_smooth(span = .15) +
  theme_minimal()
```




























