---
title: "session 4 notes"
author: "Tim Riffe"
date: "2024-07-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Get data (pasted in Doc)

```{r}
library(tidyverse)
library(readr)
library(janitor)
source("https://raw.githubusercontent.com/timriffe/BSSD2024Module2/master/02_lifetables.R")
B <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2024Module2/master/data/ES_B2014.csv") |> 
  mutate(sex = "total") |> 
  select(age, sex, births = total)

D<- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2024Module2/master/data/ES_D2014.csv") |> 
     filter(year == 2014) |> 
  select(-open_interval) |> 
  pivot_longer(female:total, 
               names_to = "sex", 
               values_to = "deaths")

P <- read_csv("https://raw.githubusercontent.com/timriffe/BSSD2024Module2/master/data/ES_P.csv.gz") |> 
  clean_names() |> 
  select(-open_interval) |> 
  filter(year== 2014) |> 
  pivot_longer(female1:total2, 
               names_to = "sex", 
               values_to = "pop") |> 
  mutate(period = parse_number(sex),
         sex = gsub('[[:digit:]]+', '', sex)) |> 
  pivot_wider(names_from = period, values_from = pop, names_prefix="pop")

ES2014 <- 
  left_join(P, D, by = join_by(year,sex, age)) |> 
  left_join(B, by = join_by(sex, age))

LT <- 
  ES2014 |> 
  mutate(exposure = (pop1 + pop2) / 2,
         mx = deaths / exposure) |> 
  select(sex,age,mx) |> 
  group_by(sex) |> 
  group_modify(~LT_tidy(data = .x)) |> 
  ungroup()
```

Check out what we have:

```{r}
ES2014 |> 
  filter(sex == "total") |> 
  mutate(exposure = (pop1 + pop2) / 2) |> 
  summarize(E = sum(exposure),
            D = sum(deaths),
            B = sum(births, na.rm = TRUE),
            G = sum(pop2) - sum(pop1),
            P1 = sum(pop1),
            P2 = sum(pop2)) |> 
  mutate(CGR = G / E,
         CBR = B / E,
         CDR = D / E,
         CNMR = (P2 - (P1 + B - D)) / E,
         CNGR = CBR - CDR)
```

Let's build intuition on how growth rates can shape population age structure, using stable populations:

```{r}

r_values <- tibble(r = seq(-.03,.03,by=.005))
library(colorspace)
LT |> 
  filter(sex == "female") |> 
  select(age, Lx) |> 
  cross_join(r_values) |> 
  arrange(r, age) |> 
  group_by(r) |> 
  mutate(Cx = Lx * exp(-r*(age+.5)),
         Cx = Cx / sum(Cx)) |> 
  ggplot(aes(x = age, y = Cx, color = r, group = r)) +
  geom_line() +
  scale_color_binned_sequential() +
  theme_minimal()
```

# Let's gather what we need in order to do a population projection

1. Calculate ASFR (births / exposure)
```{r}
expos <-
  ES2014 |> 
  filter(sex == "female") |> 
  mutate(exposure = (pop1 + pop2) / 2) |> 
  select(age, exposure)
asfr <-
  B |> 
  select(-sex) |> 
  right_join(expos, by = join_by(age)) |> 
  arrange(age) |> 
  mutate(births = if_else(is.na(births),0,births),
         asfr = births / exposure) |> 
  select(age, asfr)

# asfr |> 
#   ggplot(aes(x = age, y = asfr)) +
#   geom_line()
```

2. Calculate Survival probabilities

$$ S_x = \frac{L_{x+1}}{L_x}$$

```{r}
Sx <-
  LT |> 
  filter(sex == "female") |> 
  select(age, Lx) |> 
  mutate(Sx = lead(Lx) / Lx) |> # changed mind here; removed 0 padding
  filter(!is.na(Sx)) # NEW
```

These are what is used to survive the members of one age to the next age one year in the future.

```{r}
L <-
  Sx |> 
  select(-Lx) |> 
  mutate(age_to = age + 1) |> 
  pivot_wider(names_from = age,
              values_from = Sx,
              values_fill = 0) |> 
  column_to_rownames("age_to") |> 
  as.matrix()
```

Now let's pad this matrix with a row of 0s on top and a column of 0s on the right:

```{r}
Les <-
  0 |> 
  rbind(L) |> 
  cbind(0)
# cbind(rbind(0, L), 0)
```

If you want to let the open age group linger around (not permanently) then put this in this in the lower right-hand corner:

$$ \frac{T_\omega}{T_{\omega - 1}}$$ 
```{r, eval = FALSE}
Tx_closeout <-
LT |> 
  filter(sex == "female",
  age >= 109) |> 
  pull(Tx)

Les[nrow(Les), ncol(Les)] <- Tx_closeout[2] / Tx_closeout[1]
```

# Fertility

Now, we have to account for fertility somehow, noting:
1. age groups need to be blended
2. we need to discount for the survival (in the age group) of mothers
3. we need to discount for girl births only
4. we need to discount for the survival of infants to age 0 on Dec 31.

Where does this come from?
```{r}
Fx  <- asfr$asfr
PF  <- 1 - .53 # Pawel's guess or preference
.4886 # Ken Wachter's favorite value for PF
SRB <- 1.05 # common perception
PF  <- 1 / (1 + SRB)
```
PF turns out to be quite important for this family of model, remember that

```{r}
L0 <- LT |> 
  filter(sex == "female",
         age == 0) |> 
  pull(Lx)

sx <-
  Sx |> 
  pull(Sx)
sx <- c(sx,0)

top_row <- L0 * PF * (Fx + lead(Fx,default = 0)) / 2 * sx
```

# Now impute this row into the Leslie matrix

First index is rows, second index is columns
```{r}
Les[1, ] <- top_row
```

How to project 1 year using this matrix
```{r}
N <- 
  P |> 
  filter(sex == "female") |> 
  pull(pop2)

# plot(0:110, N, type = 'l')
# lines(0:110, c(Les %*% N), col = "blue")
```

# Let's do a longer-term population projection

To carry the population over longer time horizons, you can take matrix powers. See below: we can keep multiplying the Leslie matrix into the output of the previous multiplication h times, or we could just raise the Leslie matrix to the hth power. Using a matrix exponential. 
```{r, eval = FALSE}
Les %*% (Les %*% (Les %*% N))

# install.packages("expm")
library(expm)
(Les %^% 3) %*% N
```

To do a longer run and filled-out population projection, you can set it up like so:
```{r}
horizon <- tibble(h = 0:200)

proj <-
  P |> 
  filter(sex == "female") |> 
  select(year, age, p0 = pop2) |> 
  cross_join(horizon) |> 
  arrange(h, age) |> 
  group_by(h) |> 
  mutate(pop = c((Les %^% h[1]) %*% p0 ))
```










